A geometric relationship of $F_2, F_3$ and $F_4$-statistics with Principal Component Analysis

### 2022-02-07 fstats_pca_lewontin.tex
revision3 part 1

\title{A geometric relationship of $F_2, F_3$ and $F_4$-statistics with Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Principal Component Analysis (PCA) and $F$-statistics \textit{sensu} Patterson are two of the most widely used population genetic tools to study human genetic variation. Here, I derive explicit connections between the two approaches and  show that these two methods are closely related. $F$-statistics have a simple geometrical interpretation in the context of PCA, and orthogonal projections are a key concept to establish this link. I show that negative $F_3$ corresponds to a circle on a PCA plot, and that $F_4$ is related to an angle measurement. I illustrate my results on two examples, one of Western Eurasian, and one of global human diversity. In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and that PCA-plots are effective at predicting $F$-statistics. My results extend $F$-statistics to more continuous population models, moving towards a more complete descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

@@

To make sense of outgroup-$F_3$-statistics in the PCA context, I use the association of $F_3$-statistics to projections (Equation \ref{eq:proj}):
On a PCA-plot, we can visualize this $F_3$-statistic as the projection of the vector $X_i - X_O$ onto $X_U - X_O$: 
$$\vectorproj[X_U - X_O]{X_i - X_O} =F_3(X_O; X_U, X_i) \frac{X_U - X_O}{F_2(X_O; X_U)}.$$

@@

The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure. On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs. Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs as there local differentiation, which is typically reflected on higher PCs, becomes important.

@@

\subsection{Assumptions}
The other cause for disagreements between $F$-statistics and PCA are differences in assumptions. The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized. In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted, and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies. In contrast, most data analyses have to grapple with missing data, SNPs are often weighted according to their allele frequencies, and observed, individual-level genotypes are used as the basis of PCA.

\subsubsection{Missing data}
The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing missing data have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}. Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}. In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation. These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}. One alternative suggested by the theory developed here could be to first perform a PCA to impute missing data, and then calculate $F$-statistics using this imputed data set. Using this approach, the accuracy of $F$-statistics including low-coverage samples would be greatly increased, as they can ``borrow'' information from closely related samples.

@@

\subsubsection{Individual vs. population-based analyses}
The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations. This is not necessary; population-based PCA has been the default in the past \citep{cavalli-sforza1994}, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}. Thus, more commonly the choice between individual-based and population-based approaches is in the goal of an analysis. Grouping individuals into populations introduces a major assumption, that can e.g. be justified using an individual-based PCA. In particular, since $F$-statistics assume individuals are randomly drawn from a population, they should form tight clusters on an individual-based PCA-plot, otherwise population substructure becomes a possible alternative model for negative $F_3$-statistics and non-zero $F_4$-statistics \citep{peter2016}.

>>>>

\title{A geometric relationship of $F_2, F_3$ and $F_4$-statistics with Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Principal Component Analysis (PCA) and $F$-statistics \textit{sensu} Patterson are two of the most widely used population genetic tools to study human genetic variation. Here, I derive explicit connections between the two approaches and  show that these two methods are closely related. $F$-statistics have a simple geometrical interpretation in the context of PCA, and orthogonal projections are a key concept to establish this link. 
I show that for any pair of populations, any population that is admixed as determined by an $F_3$-statistic will lie inside a circle on a PCA plot. Furthermore, the $F_4$-statistic is closely related to an angle measurement, and will be zero if the differences between pairs of populations intersect at a right angle in PCA space. 
I illustrate my results on two examples, one of Western Eurasian, and one of global human diversity. In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and that PCA-plots are effective at predicting $F$-statistics. Thus, while F-statistics are commonly understood in terms of discrete populations, the geometric perspective illustrates that they can be viewed in a framework of populations that vary in a more continuous manner.
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

@@

To make sense of outgroup-$F_3$-statistics in the PCA context, I use the association of $F_3$-statistics to projections (Equation \ref{eq:proj}):
On a PCA-plot, we can visualize this $F_3$-statistic as the projection of the vector $X_i - X_O$ onto $X_U - X_O$: 
$$\vectorproj[X_U - X_O]{(X_i - X_O)} =F_3(X_O; X_U, X_i) \frac{X_U - X_O}{F_2(X_O; X_U)}.$$

@@

The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure. On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs \citep{cavalli-sforza1975}. Thus, I would expect that approximating $F$-statistics with few PCs works well whenever local variation is unimportant, i.e. when calculating $F$-statistics far apart in PCA space. On the flip side, populations close-by on a PCA-plot are often seperated on higher PCs, and thus whenever close-by populations are included in the same $F$-statistics, these higher PCs will be required for good approximations.

@@

\subsection{Assumptions}
In addition to the selection of PCs, the other cause for disagreements between $F$-statistics and PCA are differences in assumptions. The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized. In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted, and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies. In contrast, most data analyses have to grapple with missing data, SNPs are often weighted according to their allele frequencies, and observed, individual-level genotypes are used as the basis of PCA.

\subsubsection{Missing data}
The matrix decompositions underlying PCA assume complete data, and thus cannot be used when some data is missing \citep{jollife2013}.  As missing data is a very common practical problem, there is a large number of algorithms for imputing missing data. The simplest approach is to replace missing data with zeros \citep[as implemented e.g. in][]{patterson2006}, but more sophisticated algorithms exist to ``learn'' the missing values from surrounding data \citep[e.g.][]{hastie2015, meisner2018, meisner2021}. In contrast, missing data in $F$-statistics is most commonly handled by estimating a standard error by resampling along the genome \citep{patterson2012}, and so missing data results in larger standard error.

These strategies are distinct, and reflect the original purposes of the approaches. For statistical tests based on $F$-statistics, we wish to isolate a set of three or four populations and get our best guess based on just that subset of data. In contrast methods for PCA can leverage the additional individuals, and thus will likely result in more accurate estimates.

However, the way we handle missing data is not tied into the method. For example, we could evaluate the robustness of a PCA by resampling data. Similarly, the theory developed here suggests that we could obtain accurate $F$-statistics with missing data by first performing a PCA using a method that handles missing data, and then calculate $F$-statistics from these PCs.

@@

\subsubsection{Individual vs. population-based analyses}
The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations.However, population-based PCA has been the default in the past \citep{cavalli-sforza1994}, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}. Often, an individual-based PCA is used to justify grouping individuals into populations; i.e. individuals that form a tight cluster on a PCA-plot have similar relationships to everyone else in the data set, and can thus be treated as a unit of analysis. Thus, if the assumptions are satisfied, $F$-statistics for individual-based and population-based analyses are expected to be very similar. PCA, on the other hand, is strongly impacted by the number of individuals from each population \citep[e.g.][]{mcvean2009}; as each individual is weighted equally, variation related to populations with many samples will be overrepresented on the first PCs.
